---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Samriddha Ghosh"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
getwd()
library(tidyverse)
library(ggplot2)
library(agricolae)
library(lubridate)
library(formatR)


NTL_LTER<-read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

NTL_LTER$sampledate<-mdy(NTL_LTER$sampledate)
class(NTL_LTER$sampledate)


```

```{r themebuilding}
#2

mytheme<-theme_bw(base_size = 12) +
  theme(axis.text = element_text(color="black"),
        legend.position = "top", 
        legend.justification = "centre")

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: No sig relation bw temp and depth across lakes There is no difference in the mean
temperature with depth from lake to lake
Ha: There is a significant difference in the mean temperatures that vary with depth in the lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r wrangling}
#4

NTL_LTER_df<-select(NTL_LTER, lakename, year4, daynum, sampledate,depth,temperature_C)%>%
  filter(month(sampledate)==7)%>%
  drop_na(temperature_C)
NTL_LTER_df
```

```{r scatterplot}
#5
graph1<-ggplot(NTL_LTER_df, aes(x = depth, y = temperature_C, color=lakename)) + 
  geom_point() + 
  geom_smooth(method = lm, color = "black") + ylim(0, 40) +
  labs(x="Depth in metres",
       y="Temperatures in degree Celcius",
       title="Temperature vs depth")
print(graph1)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The relation between depth and temperature is found to be inversely linear. Temperature lowers as depth rises. From 10m to 15m, the temperature appears to remain steady, with the points indicating a nearly linear trend.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
NTL.regression<-lm(NTL_LTER_df$temperature_C ~ NTL_LTER_df$depth)
summary(NTL.regression)

```

 
8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The results of the model show that the change in depth can account for 73.87% of the variation in temperature. The p-value is 2.2e-16, which is less than 0.05, hence the null hypothesis was not accepted. 9501 Degrees of Freedom serve as its foundation. The temperature lowers by -1.6C for each additional 1m of depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

TempAIC <- lm(data = NTL_LTER_df, temperature_C ~ depth + year4 + daynum) 
step(TempAIC)
Tempmodel<-lm(data=NTL_LTER_df, temperature_C ~ depth + year4 + daynum)
summary(Tempmodel)

#10

TempRegression<-lm(data=subset(NTL_LTER_df, lakename=="Paul Lake"),temperature_C ~ depth + daynum)
summary(TempRegression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The best predictor variables to employ are depth and daynum, according to the findings of the AIC approach. As it was statistically insignificant, we discarded the year. The variation in depth accounts for 83.1% of the temperature variability. As it makes use of several explanatory variables, it is an improvement over the prior model, which had an explanatory power of about 73.8%.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

NTL.anova<-aov(data=NTL_LTER_df, temperature_C ~ depth + daynum + lakename)
summary(NTL.anova)

NTL.anova1 <- lm(data = NTL_LTER_df, temperature_C ~ depth + daynum + lakename) 
summary(NTL.anova1)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Except for Ward Lake and Long Lake, which have a higher p-value, there appears to be a substantial difference in the mean temperature among the other lakes in July.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot2}
#14.
graph2 <- ggplot(NTL_LTER_df, aes(x = depth, y = temperature_C, color = lakename), alpha = 0.5) + 
  geom_point() + 
  geom_smooth(method = lm, color = "black", se = FALSE) + ylim(0, 40) + facet_wrap(~lakename)+
  labs(x="Depth in metres",
       y="Temperatures in degree Celcius",
       title="Plot of temperature vs depth for different lakes")
print(graph2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TSD_turkey<-aov(data = NTL_LTER_df, temperature_C ~ lakename)
summary(TSD_turkey)
TukeyHSD(TSD_turkey)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:We can observe that the pairs Ward Lake - Peter Lake and Peter Lake - Paul Lake have a p value better than 0.1, indicating that the mean temperature of these three lakes is the same. As each set of pair-wise comparisons contains one pair with a p-value greater than 0.05, signifying no statistically significant difference in mean temperatures, it appears that no lakes have a mean temperature that is statistically unique from all other lakes.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we want to determine whether the means of two samples differ substantially from one another or not, we can apply a one-way ANOVA.

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
NTL_Crampton <- NTL_LTER_df %>%
filter(lakename == 'Crampton Lake')
NTL_Ward <- NTL_LTER_df %>%
filter(lakename == 'Ward Lake')
NTL_crampton_ward <- rbind(NTL_Crampton, NTL_Ward)


t.test(x = NTL_Crampton$temperature_C,
y = NTL_Ward$temperature_C,
alternative = "two.sided")
```

>Answer: The T score for the two-sample t-test was 1.1181 on 200.37 degrees of freedom, which is greater than 0.025 and translates to a p-value of 0.2649.
We conclude that the mean temperature difference between the two lakes in July is not statistically significant, assuming a 95% confidence interval.
