---
title: 'Assignment 5: Data Visualization'
author: "Samriddha Ghosh"
date: "Spring 2023"
output:
  word_document: default
  pdf_document: default
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options:
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 Loading packages
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

getwd() #checking directory

Litter<-read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)
Nutrient<-read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

#2 Changing the format to date
Litter$collectDate<-ymd(Litter$collectDate)
Nutrient$sampledate<-ymd(Nutrient$sampledate)
Nutrient$month<-month(Nutrient$month, label=TRUE)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3 Defining the plot by:
#a)changing the plot background
#b)Setting the position and alignment of the plot legend
#c)Changing the color of the axis texts to black
mytheme<-theme_bw(base_size = 12) +
  theme(axis.text = element_text(color="black"),
        legend.position = "top", 
        legend.justification = "centre")

theme_set(mytheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r, fig.height=5, fig.width=5}
#4  

ggplot(Nutrient, aes(x= po4, y= tp_ug, color=lakename, shape=lakename))+
  geom_point()+
  xlim(-2,50)+
  ylim(-2,160)+
  geom_smooth(method=lm, color="black", size=0.5)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 

temperature<-ggplot(Nutrient, aes(x=month, y=temperature_C))+
  geom_boxplot(aes(color = lakename))+
  xlab("Months of the year")+
  ylab("Temperatures in Celcius")
print(temperature)

TP<-ggplot(Nutrient, aes(x=month, y=tp_ug, color=lakename))+
  geom_boxplot()+
  xlab("Months of the year")+
  ylab("Total Phosphorus in ug")

print(TP)

TN<-ggplot(Nutrient, aes(x=month, y=tn_ug, color=lakename))+
  geom_boxplot()+
  xlab("Months of the year")+
  ylab("Total Nitrogen in ug")
print(TN)
```


```{r Creating combined boxplot}
Legend_less<-plot_grid(temperature + theme(legend.position = "none"),
          TP + theme(legend.position = "none"),
          TN + theme(legend.position = "none"),
          nrow=3, align='h')
print(Legend_less)

Legend<-get_legend(
  temperature)

print(plot_grid(Legend_less, Legend, nrow=2, rel_heights = c(5,0.1)))

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 
In case of plot 1 for temperature against the months of the year: The mean temperature of Paul lake is higher than Peter lake up until September after which the mean temperature of Peter lake increases upto November.
In case of plot 2 for total phosphorus in ug vs months of the year: The mean of the total phosphorus in ug for Peter lake keeps increasing with each passing month.
In case of plot 2 for total nitrogen in ug vs months of the year: The mean of the total nitrogen in ug for Peter lake keeps increasing with each passing month.


6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
Litter %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(aes(y = dryMass, x = collectDate, color = nlcdClass)) +
  geom_point()


#7

Litter %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(aes(y = dryMass, x = collectDate)) +
    geom_point()+
    facet_wrap(vars(nlcdClass),nrow=1)


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: The graph plotted in Q.7 is more effective. 
This is because now for each nlcdClass, we are automatically getting a separate plot for drymass against a date. This graph makes the data visualization and comparison easier and better.